---
title: "LexiGover_HW2"
output: word_document
date: "2026-02-02"
---
1. Read this tab-separated text file sleephealth.txt into R and call it "sleep". Present the variable names and the dataset's dimensions. (2 points)
```{r}
sleep = read.table("sleephealth.txt", header= TRUE, sep = "\t")
names(sleep)
dim(sleep)
```


2. Use the apply() function to estimate the means of age and heart rate, rounded to 3 decimal places (3 d.p.) in a single R statement. (2 points)
```{r}
round(apply(sleep[,c("Age", "Heart_Rate")], 2, mean), 3)
```


3. Use the lapply() function to estimate the means of age and heart rate in a single R statement (not required to round to 3 d.p.)  (1 point)
```{r}
lapply(sleep[, c("Age", "Heart_Rate")], mean)
```
4. Use the sapply() function to estimate the means of age and heart rate, rounded to 3 d.p., in a single R statement. What are the differences in the output for your results among parts (2), (3), and (4), other than the rounding? (2 points)
```{r}
round(sapply(sleep[, c("Age", "Heart_Rate")], mean), 3)
```
The difference is format it is displayed in. Sapply and apply display in vector format while lapply displays in list format. 


5. Use the tapply() function to estimate the mean of heart rate by Sleep_disorder group. In your R code, ensure the output is rounded to 2 d.p. (1.5 points)
```{r}
round(tapply(sleep$Heart_Rate, sleep$Sleep_disorder, FUN = mean), 2)
```


6. Create a new indicator factor variable, duration_cat.f, in the sleep data set that indicates whether each participant has a healthy sleep duration. (duration_cat.f = 0 when Sleep_duration < 7 hours; otherwise duration_cat.f = 1). Label the factor levels as follows: "short sleep duration" for individuals with a sleep duration of less than 7 hours and "normal/healthy sleep duration " for those with a sleep duration of 7 hours or more, using the labels option. Present a frequency table of duration_cat.f. (2.5 points)
```{r}

sleep$duration_cat.f = ifelse (sleep$Sleep_Duration < 7,0,1)

sleep$duration_cat.f = factor(sleep$duration_cat.f, levels = c(0,1), labels = c("short sleep duration", "normal/healthy sleep duration"))
table(sleep$duration_cat.f)
```


7. Create a new categorical factor variable, duration_4cat, in the sleep dataset that categorizes the sleep duration into four categories with the following labels: "<6", "[6,7)", "[7,8)", ">=8". The intervals should be left-closed and right-open, meaning that, for example, the "[6,7]" category includes values â‰Ą6 and <7. Create a frequency table of the duration_4cat variable. (2.5 points)
```{r}
sleep$duration_4cat = ifelse(sleep$Sleep_Duration < 6,0, 
                        ifelse(sleep$Sleep_Duration<7,1,
                       ifelse(sleep$Sleep_Duration<8,2,3)))

sleep$duration_4cat <- factor(sleep$duration_4cat,
                              levels = c(0,1,2,3),
                              labels = c("<6", "[6,7)", "[7,8)", ">=8"))

table(sleep$duration_4cat)
```


8. Create a new data frame called SleepDurationdata with only two variables: Sleep_Duration in the first column and duration_4cat in the second column. Use the head() function to show the first six rows of the new data frame. What are the dimensions of Sleep_Durationdata (no. of rows by the no. of columns)? Create a frequency table of duration_4cat with Sleep_Durationdata. (2.5 points)
```{r}
SleepDurationdata = sleep[,c("Sleep_Duration", "duration_4cat")]
head(SleepDurationdata)
dim(SleepDurationdata)

table(SleepDurationdata$duration_4cat)
```

9. Export your the SleepDurationdata data frame and save it as a csv file called "mydata.csv". (1 point)
```{r}
write.table(SleepDurationdata, "mydata.csv")
```


10. Cross-tabulate the variables BMIcat and gender in the sleep dataset. Also, present a table of proportions of BMIcat for each level of gender. From these tables, how many males are there with a normal BMI? Among the males in the sample, what percentage have a normal BMI? (3 points)
```{r}
table(sleep$BMIcat, sleep$Gender)
prop.table(table(sleep$BMIcat, sleep$Gender), 2) *100
```
There are 138 males with a normal BMI. 73.02% of males have a normal BMI. 




