---
title: "LexiGover_HW3"
output: word_document
date: "2026-02-09"
---

a) Read the two text files into R and assign them to two objects called baseline and followup respectively. Present the variable names and dimensions of each dataset. (2 points)
```{r}
baseline = read.table("baseline_vision.txt", header=TRUE, sep=" ")
followup = read.table("followup_vision.txt", header=TRUE, sep=" ")

names(baseline)
names(followup)

dim(baseline)
dim(followup)
```


b) Merge the two data frames into a single data frame called vision_data by the patient’s id and eye. After merging, remove any rows with missing values using vision_data = na.omit(vision_data). What are the variable names and dimensions of vision_data? (2 points)
```{r}
vision_data = merge(baseline, followup, by= c("id", "eye"))

vision_data = na.omit(vision_data)

names(vision_data)
dim(vision_data)
```
Variable names: id, eye, laser, trt, age, risk, vision_baseline, vision_6mo

The dimensions are 389 rows (observations) x 8 columns (variables).


c) Rename the variable trt to treatment and risk to risk_score in the vision_data dataset. Present the updated variable names. (2 points)
```{r}
library(reshape)

vision_data = rename(vision_data, c("trt" = "treatment", "risk" = "risk_score"))
names(vision_data)
head(vision_data)
```


d) Drop the variables laser and risk_score from the vision_data dataset. Present the updated variable names. (2 points)
```{r}
vision_data = subset(vision_data, select = -c(laser, risk_score))

names(vision_data)
head(vision_data)
```


e) Create a new factor variable eye.f in the vision_data dataset with labels “Left Eye” and “Right Eye”. Present a frequency table for eye.f. (2 points)
```{r}
vision_data$eye.f = factor(vision_data$eye, levels = c("left", "right"), labels = c("Left Eye", "Right Eye"))

table(vision_data$eye.f)
```

f) Calculate the change in vision score for each observation by subtracting the baseline vision score from the 6-month vision score. Call this new variable vision_change and save it in the vision_data dataset. A negative value indicates vision deterioration. What are the mean and standard deviation of vision_change? Round to 2 decimal places. (2 points)
```{r}
vision_data$vision_change = (vision_data$vision_6mo - vision_data$vision_baseline)

round(mean(vision_data$vision_change),2)
round(sd(vision_data$vision_change),2)
```

The mean is -3.07 and the standard deviation is 1.03.


g) Create a new variable age_group and save it in the vision_data dataset. This variable should categorize patients into three groups: (3 points)
```{r}
vision_data$age_group = rep(0, length(vision_data$age)) 
vision_data$age_group[vision_data$age < 20] = "Young"
vision_data$age_group[vision_data$age >= 20 & vision_data$age <= 40] = "Middle"
vision_data$age_group[vision_data$age >40] = "Senior"

vision_data$age_group = factor(vision_data$age_group)

table(vision_data$age_group)
```


h) Subset the vision_data dataset to include only observations where vision_change is negative (i.e., vision deteriorated). Call this subset deteriorated. What are the dimensions of deteriorated? What percentage of the total observations had vision deterioration? (2 points)
```{r}
deteriorated = subset(vision_data, vision_change < 0)

dim(deteriorated)

percent_deteriorate = nrow(deteriorated) / nrow(vision_data) *100

percent_deteriorate
```
The dimensions are 387 rows (observations) x 9 columns (variables). 99.49% of the total observations has vision deteriorated.


i) Using the deteriorated dataset from part (h), create a cross-tabulation of age_group and treatment. Among patients who experienced vision deterioration, which age group had the highest number of treated eyes? (3 points)
```{r}
table(deteriorated$age_group, deteriorated$treatment)
```
The young had the highest number of treated eyes among the patients who experienced vision deterioration.
